---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dml.sensemakr

<!-- badges: start -->
<!-- badges: end -->

`dml.sensemakr` implements a general suite of sensitivity analysis tools for Causal Machine Learning as discussed in [Chernozhukov, V., Cinelli, C., Newey, W., Sharma A., and Syrgkanis, V. (2021). "Long Story Short: Omitted Variable Bias in Causal Machine Learning."](https://www.nber.org/papers/w30302)


## Development version

To install the development version on GitHub make sure you have the package devtools installed.

```{r, eval = FALSE}
# install.packages("devtools") 
devtools::install_github("carloscinelli/dml.sensemakr")
```
# CRAN

CRAN version coming soon.


# Details

For theoretical details of the method [please see the working paper.](https://www.nber.org/papers/w30302) For a primer on Debiased Machine Learning, 
[please check Chernozukohv et al (2018).](https://academic.oup.com/ectj/article/21/1/C1/5056401) 

Some presentations that may be useful:

- [Carlos' presentation at the ICLR.](http://interactivecausallearning.com/Carlos_Cinelli.html)
- [Victor's tutorial at the Chamberlain Seminar.](https://www.youtube.com/watch?v=PQtYqKfxH_I)

# Quick Example


## Fit DML model

### Load Data
```{r example}
# loads package
library(dml.sensemakr)

## loads data
data("pension")

# set the outcome
y <- pension$net_tfa  # net total financial assets

# set the treatment
d <- pension$e401    # 401K eligibility

# set the covariates (a matrix)
x <- model.matrix(~ -1 + age + inc  + educ+ fsize + marr + twoearn + pira + hown, data = pension)

## compute income quartiles for group ATE.
g1 <- cut(x[,"inc"], quantile(x[,"inc"], c(0, 0.25,.5,.75,1), na.rm = TRUE), 
          labels = c("q1", "q2", "q3", "q4"), include.lowest = T)
```

### Use DML to estimate the ATE

```{r}
# run DML (nonparametric model)
## 2 folds (change as needed)
## 1 repetition (change as needed)
dml.401k <- dml(y, d, x, model = "npm", groups = g1, cf.folds = 2, cf.reps = 1)
```

### Explore results
```{r}
# summary of results with median method (default)
summary(dml.401k, combine.method = "median")
```

### extract coefs, se, confidence intervals
```{r}
# coef median method (default)
coef(dml.401k, combine.method = "median")

# se median method (default)
se(dml.401k, combine.method = "median")

# confint median method
confint(dml.401k, combine.method = "median")
```

<!-- ### Coefficients Plot -->

<!-- ```{r} -->
<!-- plot(dml.401k) -->
<!-- ``` -->

<!-- ### You can add groups after the model is fit -->

<!-- ```{r} -->
<!-- ## compute GATE by married -->
<!-- g2 <- ifelse(pension$marr, "married", "not.married") -->
<!-- dml.401k.g2 <- dml_gate(dml.fit = dml.401k, groups = g2) -->
<!-- summary(dml.401k.g2) -->

<!-- coef(dml.401k.g2) -->
<!-- confint(dml.401k.g2) -->
<!-- plot(dml.401k.g2) -->
<!-- ``` -->

## Sensitivity Analysis

### Robustness Values

```{r}
robustness_value(dml.401k, alpha = 0.05)
```

### Confidence Bounds

```{r}
confidence_bounds(dml.401k, r2ya.dx = 0.03, r2.rr = 0.04, level = 0.95)
```

### Contour Plots

```{r}
ovb_contour_plot(dml.401k, r2ya.dx = 0.03, r2.rr = 0.04, bound.label = "Max Match (3x years)")
```


<!-- ### Further results -->

<!-- ```{r} -->
<!-- bounds.401k <- dml_bounds(dml.401k, r2ya.dx = 0.03, r2.rr = 0.04) -->
<!-- bounds.401k -->
<!-- ``` -->

<!-- ```{r} -->
<!-- coef(bounds.401k) -->
<!-- se(bounds.401k) -->
<!-- confint(bounds.401k) -->
<!-- ``` -->

