---
title: "Replication: Long Story Short (RESTAT 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{restat-replication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  dpi = 300,
  fig.width = 12,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(dml.sensemakr)
library(ggplot2)
```

This vignette replicates the main empirical results of [Chernozhukov, V., Cinelli, C., Newey, W., Sharma A., and Syrgkanis, V. (2026). "Long Story Short: Omitted Variable Bias in Causal Machine Learning." *Review of Economics and Statistics*.](https://doi.org/10.1162/REST.a.1705)

The application examines the effect of 401(k) eligibility on net financial assets using the `pension` dataset. We show both the Partially Linear Model (PLM) and Nonparametric Model (NPM) analyses.

# Load Data

```{r}
library(dml.sensemakr)

## loads data
data("pension")
y <- pension$net_tfa # net total financial assets
d <- pension$e401    # 401K eligibility
x <- model.matrix(~ -1 + age + inc + educ + fsize + marr + twoearn + pira + hown, data = pension)
```

# Naive Estimate

```{r}
mean(y[d==1]) - mean(y[d==0])
```


# Partially Linear Model

## DML Estimation

```{r plm-fit}
set.seed(1)
dml.401k.plm <- dml(y, d, x, model = "plm", cf.folds = 5, cf.reps = 5)
summary(dml.401k.plm)
```

## Sensitivity Analysis

```{r plm-sensitivity}
sens.401k <- sensemakr(dml.401k.plm, cf.y = 0.04, cf.d = 0.03,
                       bound_label = "Max match (3 years)")
summary(sens.401k)
```

## Contour Plots

```{r plm-contour, fig.height=6}
oldpar <- par(mfrow = c(1,2))
plot(sens.401k, which.bound = "lwr", lim.x = 0.18, col.contour = "blue")
plot(sens.401k, which.bound = "upr", lim.x = 0.18, col.contour = "blue")
par(oldpar)
```

## Benchmarks

```{r plm-bench}
bench.plm <- dml_benchmark(dml.401k.plm, benchmark_covariates = c("inc", "pira", "twoearn"), seed = 2)
bench.plm
```

## Contour Plots with Benchmarks

```{r plm-contour-bench, fig.height=6}
# sensitivity analysis (no benchmark_covariates — we overlay manually for fine control)
sens.401k <- sensemakr(dml.401k.plm, cf.y = NULL)

# 1/4 x Income benchmark (only income)
bench.inc <- dml_benchmark(dml.401k.plm, benchmark_covariates = "inc", seed = 2)

oldpar <- par(mfrow = c(1,2))

# --- PLM lower bound, rho2 = 1 ---
plot(sens.401k, which.bound = "lwr", lim.x = 0.16, col.contour = "blue",
     label.bump.x = 0.015, label.bump.y = 0.003)
add_bound_to_contour(dml.401k.plm, cf.y = 0.04, cf.d = 0.03,
                     bound_label = "Max Match",
                     label.bump.x = 0.015, label.bump.y = 0.003,
                     point.pch = 25, point.bg = "red")
add_bound_to_contour(bench.plm, which.bound = "lwr",
                     bound_label = c("1 x Income", "1 x Part. in IRA", "1 x Two Earners"),
                     label.bump.x = c(0.01, 0.01, 0.02),
                     label.bump.y = c(0.009, 0.009, -0.002))
add_bound_to_contour(bench.inc, kd = 1/4, which.bound = "lwr",
                     bound_label = "1/4 x Income",
                     label.bump.x = 0.015, label.bump.y = -0.005)

# --- PLM lower bound, rho2 = (1/2)^2 ---
rho2 <- (1/2)^2
sens.401k.rho2 <- sensemakr(dml.401k.plm, cf.y = NULL, rho2 = rho2)
plot(sens.401k.rho2, which.bound = "lwr", lim.x = 0.16, col.contour = "blue",
     label.bump.x = 0.015, label.bump.y = 0.003)
add_bound_to_contour(dml.401k.plm, cf.y = 0.04, cf.d = 0.03, rho2 = rho2,
                     bound_label = "Max Match",
                     label.bump.x = 0.015, label.bump.y = 0.003,
                     point.pch = 25, point.bg = "red")
add_bound_to_contour(bench.plm, which.bound = "lwr", rho2 = rho2,
                     bound_label = c("1 x Income", "1 x Part. in IRA", "1 x Two Earners"),
                     label.bump.x = c(0.01, 0.01, 0.02),
                     label.bump.y = c(0.009, 0.009, -0.002))
add_bound_to_contour(bench.inc, kd = 1/4, which.bound = "lwr", rho2 = rho2,
                     bound_label = "1/4 x Income",
                     label.bump.x = 0.015, label.bump.y = -0.005)

par(oldpar)
```


# Nonparametric Model

## DML Estimation with GATE

```{r npm-fit}
g1 <- cut(x[,"inc"], quantile(x[,"inc"], c(0, 0.25, .5, .75, 1), na.rm = TRUE),
          labels = c("q1", "q2", "q3", "q4"), include.lowest = TRUE)

set.seed(1)
dml.401k.npm <- dml(y, d, x, groups = g1, model = "npm", cf.folds = 5, cf.reps = 5)
summary(dml.401k.npm)
```

## GATE Plot under Conditional Ignorability

```{r npm-gate, fig.width=8, fig.height=5}
library(ggplot2)

group.names <- paste0("gate.q", 1:4)
df   <- data.frame(groups = 1:4, estimate = coef(dml.401k.npm)[group.names])
cis  <- confint(dml.401k.npm)[group.names, ]
cis  <- setNames(as.data.frame(cis), c("lwr.ci", "upr.ci"))
df   <- cbind(df, cis)

ggplot(df, aes(x = groups, y = estimate)) + geom_line() +
  geom_ribbon(aes(ymin = lwr.ci, ymax = upr.ci), alpha = 0.1, col = "blue", fill = "blue") +
  theme_bw() +
  xlab("Income Groups by Quartiles") +
  ylab("ATE")
```

## Sensitivity Analysis

```{r npm-sensitivity}
sens.401k.npm <- sensemakr(dml.401k.npm, cf.y = 0.04, cf.d = 0.03,
                            bound_label = "Max match (3 years)")
summary(sens.401k.npm)
```

## Contour Plots

```{r npm-contour, fig.height=6}
oldpar <- par(mfrow = c(1,2))
plot(sens.401k.npm, which.bound = "lwr", lim.x = 0.18, col.contour = "blue")
plot(sens.401k.npm, which.bound = "upr", lim.x = 0.18, col.contour = "blue")
par(oldpar)
```

## GATE Bounds Plot

```{r npm-gate-bounds, fig.width=8, fig.height=5}
bds   <- confidence_bounds(dml.401k.npm, cf.y = 0.04, cf.d = 0.03, level = 0)
bds   <- setNames(as.data.frame(bds), c("lwr.bound", "upr.bound"))
cbds  <- confidence_bounds(dml.401k.npm, cf.y = 0.04, cf.d = 0.03, level = .95)
cbds  <- setNames(as.data.frame(cbds), c("lwr.cbound", "upr.cbound"))
df2   <- cbind(df, bds[-1, ], cbds[-1, ])

ggplot(df2, aes(x = groups, y = estimate)) + geom_line() +
  geom_ribbon(aes(ymin = lwr.bound, ymax = upr.bound),   alpha = 0.1, col = "red", fill = "red") +
  geom_ribbon(aes(ymin = lwr.cbound, ymax = upr.cbound), alpha = 0.1, col = "blue", fill = "blue") +
  theme_bw() +
  xlab("Income Groups by Quartiles") +
  ylab("ATE")
```

## NPM Benchmarks

```{r npm-bench}
bench.npm <- dml_benchmark(dml.401k.npm, benchmark_covariates = c("inc", "pira", "twoearn"), seed = 2)
bench.npm
```

## NPM Contour Plots with Benchmarks

```{r npm-contour-bench, fig.height=6}
# sensitivity analysis (no benchmark_covariates — manual overlay)
sens.npm <- sensemakr(dml.401k.npm, cf.y = NULL)

# 1/4 x Income benchmark (only income)
bench.npm.inc <- dml_benchmark(dml.401k.npm, benchmark_covariates = "inc", seed = 2)

oldpar <- par(mfrow = c(1,2))

# --- NPM lower bound, rho2 = 1 ---
plot(sens.npm, which.bound = "lwr", lim.x = 0.16, col.contour = "blue",
     label.bump.x = 0.015, label.bump.y = 0.003)
add_bound_to_contour(dml.401k.npm, cf.y = 0.04, cf.d = 0.03,
                     bound_label = "Max Match",
                     label.bump.x = 0.015, label.bump.y = 0.003,
                     point.pch = 25, point.bg = "red")
add_bound_to_contour(bench.npm, which.bound = "lwr",
                     bound_label = c("1 x Income", "1 x Part. in IRA", "1 x Two Earners"),
                     label.bump.x = c(-0.01, 0.005, 0.02),
                     label.bump.y = c(0.009, 0.009, 0.005))
add_bound_to_contour(bench.npm.inc, kd = 1/4, which.bound = "lwr",
                     bound_label = "1/4 x Income",
                     label.bump.x = 0.015, label.bump.y = 0.00)

# --- NPM lower bound, rho2 = (1/2)^2 ---
rho2 <- (1/2)^2
sens.npm.rho2 <- sensemakr(dml.401k.npm, cf.y = NULL, rho2 = rho2)
plot(sens.npm.rho2, which.bound = "lwr", lim.x = 0.16, col.contour = "blue",
     label.bump.x = 0.015, label.bump.y = 0.003)
add_bound_to_contour(dml.401k.npm, cf.y = 0.04, cf.d = 0.03, rho2 = rho2,
                     bound_label = "Max Match",
                     label.bump.x = 0.015, label.bump.y = 0.003,
                     point.pch = 25, point.bg = "red")
add_bound_to_contour(bench.npm, which.bound = "lwr", rho2 = rho2,
                     bound_label = c("1 x Income", "1 x Part. in IRA", "1 x Two Earners"),
                     label.bump.x = c(-0.01, 0.01, 0.02),
                     label.bump.y = c(0.009, 0.009, 0.005))
add_bound_to_contour(bench.npm.inc, kd = 1/4, which.bound = "lwr", rho2 = rho2,
                     bound_label = "1/4 x Income",
                     label.bump.x = 0.015, label.bump.y = 0.00)

par(oldpar)
```
