---
title: "Replication: Long Story Short (RESTAT 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{restat-replication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  dpi = 300,
  fig.width = 12,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(dml.sensemakr)
library(ggplot2)
```

This vignette replicates the main empirical results of [Chernozhukov, V., Cinelli, C., Newey, W., Sharma A., and Syrgkanis, V. (2026). "Long Story Short: Omitted Variable Bias in Causal Machine Learning." *Review of Economics and Statistics*.](https://doi.org/10.1162/REST.a.1705)

The application examines the effect of 401(k) eligibility on net financial assets using the `pension` dataset. We show both the Partially Linear Model (PLM) and Nonparametric Model (NPM) analyses.

# Load Data

```{r}
library(dml.sensemakr)

## loads data
data("pension")
y <- pension$net_tfa # net total financial assets
d <- pension$e401    # 401K eligibility
x <- model.matrix(~ -1 + age + inc + educ + fsize + marr + twoearn + pira + hown, data = pension)
```

# Naive Estimate

```{r}
mean(y[d==1]) - mean(y[d==0])
```


# Partially Linear Model

## DML Estimation

```{r plm-fit}
# run DML with partially linear model
set.seed(1)
dml.plm <- dml(y, d, x, model = "plm", cf.folds = 5, cf.reps = 5)

# results under conditional ignorability
summary(dml.plm)
```


## Sensitivity Analysis with Benchmarks

We use `sensemakr()` with `benchmark_covariates` to simultaneously compute sensitivity bounds and benchmark covariate strengths. The benchmarks `twoearn` (two earners), `pira` (participation in IRA), and `inc` (income) are the same ones used in the paper.

```{r plm-sensitivity}
sens.plm <- sensemakr(dml.plm, cf.y = 0.04, cf.d = 0.03,
                      benchmark_covariates = c("twoearn", "pira", "inc"))
summary(sens.plm)
```

## Contour Plots (Figure 1)

Contour plots show how the lower and upper confidence bounds change as a function of the sensitivity parameters. Benchmark covariates are overlaid automatically.

```{r plm-contour, fig.height=6}
oldpar <- par(mfrow = c(1,2))
plot(sens.plm, which.bound = "lwr", lim.x = 0.18)
plot(sens.plm, which.bound = "upr", lim.x = 0.18)
par(oldpar)
```

## Contour Plots with Multipliers (Figure 2)

We can show multiples of each benchmark's strength using `kd` (and `ky`). Here we show both full strength (`1x`) and a quarter of it (`1/4x`).

```{r plm-contour-k, fig.height=6}
sens.plm.k <- sensemakr(dml.plm, cf.y = 0.04, cf.d = 0.03,
                         benchmark_covariates = c("twoearn", "pira", "inc"),
                         kd = c(1, 1/4))

oldpar <- par(mfrow = c(1,2))
plot(sens.plm.k, which.bound = "lwr", lim.x = 0.18)
plot(sens.plm.k, which.bound = "upr", lim.x = 0.18)
par(oldpar)
```

## Relaxing $\rho^2$ (Figure 3)

Setting `rho2 < 1` relaxes the worst-case assumption on the correlation between the bias components. With $\rho^2 = (1/2)^2 = 0.25$, bounds become tighter.

```{r plm-contour-rho2, fig.height=6}
sens.plm.rho <- sensemakr(dml.plm, cf.y = 0.04, cf.d = 0.03,
                           benchmark_covariates = c("twoearn", "pira", "inc"),
                           rho2 = (1/2)^2)

oldpar <- par(mfrow = c(1,2))
plot(sens.plm.rho, which.bound = "lwr", lim.x = 0.18)
plot(sens.plm.rho, which.bound = "upr", lim.x = 0.18)
par(oldpar)
```

## Manual Benchmark Overlay

You can also compute benchmarks separately with `dml_benchmark()` and add them to a contour plot step by step using `add_bound_to_contour()`:

```{r plm-manual, fig.height=6}
bench.plm <- dml_benchmark(dml.plm, benchmark_covariates = c("twoearn", "pira", "inc"))
bench.plm

ovb_contour_plot(dml.plm, which.bound = "lwr", lim.x = 0.18)
add_bound_to_contour(bench.plm, kd = c(1, 1/4))
```


# Nonparametric Model

## DML Estimation with GATE

```{r npm-fit}
# compute income quartiles for GATE analysis
g1 <- cut(x[,"inc"], quantile(x[,"inc"], c(0, 0.25, .5, .75, 1), na.rm = TRUE),
          labels = c("q1", "q2", "q3", "q4"), include.lowest = TRUE)

# run DML with nonparametric model
set.seed(1)
dml.npm <- dml(y, d, x, groups = g1, model = "npm", cf.folds = 5, cf.reps = 5)

# results under conditional ignorability
summary(dml.npm)
```

## GATE Plot under Conditional Ignorability

```{r npm-gate, fig.width=8, fig.height=5}
library(ggplot2)

group.names <- paste0("gate.q", 1:4)
df   <- data.frame(groups = 1:4, estimate = coef(dml.npm)[group.names])
cis  <- confint(dml.npm)[group.names, ]
cis  <- setNames(as.data.frame(cis), c("lwr.ci", "upr.ci"))
df   <- cbind(df, cis)

ggplot(df, aes(x = groups, y = estimate)) + geom_line() +
  geom_ribbon(aes(ymin = lwr.ci, ymax = upr.ci), alpha = 0.1, col = "blue", fill = "blue") +
  theme_bw() +
  xlab("Income Groups by Quartiles") +
  ylab("ATE")
```

## Sensitivity Analysis with Benchmarks

```{r npm-sensitivity}
sens.npm <- sensemakr(dml.npm, cf.y = 0.04, cf.d = 0.03,
                      benchmark_covariates = c("twoearn", "pira", "inc"))
summary(sens.npm)
```

## Contour Plots with Benchmarks

```{r npm-contour, fig.height=6}
oldpar <- par(mfrow = c(1,2))
plot(sens.npm, which.bound = "lwr", lim.x = 0.18)
plot(sens.npm, which.bound = "upr", lim.x = 0.18)
par(oldpar)
```

## GATE Bounds Plot

The GATE bounds plot shows how the group-level effects change under confounding. The red ribbon shows the bias-adjusted bounds (point estimates); the blue ribbon shows the confidence bounds.

```{r npm-gate-bounds, fig.width=8, fig.height=5}
bds   <- confidence_bounds(dml.npm, cf.y = 0.04, cf.d = 0.03, level = 0)
bds   <- setNames(as.data.frame(bds), c("lwr.bound", "upr.bound"))
cbds  <- confidence_bounds(dml.npm, cf.y = 0.04, cf.d = 0.03, level = .95)
cbds  <- setNames(as.data.frame(cbds), c("lwr.cbound", "upr.cbound"))
df2   <- cbind(df, bds[-1, ], cbds[-1, ])

ggplot(df2, aes(x = groups, y = estimate)) + geom_line() +
  geom_ribbon(aes(ymin = lwr.bound, ymax = upr.bound),   alpha = 0.1, col = "red", fill = "red") +
  geom_ribbon(aes(ymin = lwr.cbound, ymax = upr.cbound), alpha = 0.1, col = "blue", fill = "blue") +
  theme_bw() +
  xlab("Income Groups by Quartiles") +
  ylab("ATE")
```

## NPM Benchmarks

```{r npm-bench}
bench.npm <- dml_benchmark(dml.npm, benchmark_covariates = c("twoearn", "pira", "inc"))
bench.npm
```
